---
title: 'MSD 2019 Final Project'
subtitle: 'A replication and extension of Wage disparity and team productivity: evidence from Major League Baseball by Craig A. Depken II, 1999'
author: 'Forrest Hofmann (fhh2112) & Sagar Lal (sl3946)'
date: '`r Sys.time()`'
output:
  pdf_document:
    toc: yes
    toc_depth: 3
---

```{r setup, include=FALSE}
library(here)
library(scales)
library(tidyverse)
library(ineq)

theme_set(theme_bw())

knitr::opts_chunk$set(echo = TRUE)
```

# Introduction

## Problem Description

## Motivation

## Data Source


# Reproduction

## Reproduction Code

```{r load-data, message = FALSE}
teams <- read_csv(here('data/teams.csv'))
salaries <- read_csv(here('data/salaries.csv'))
```

```{r clean-data, message = FALSE}
teams$WSWin <- as.logical(teams$WSWin == 'Y')
teams <- teams %>%
  filter(1985 <= yearID & yearID <= 2016) %>%
  mutate(winPercentage = W / (W + L) * 1000) %>%
    filter(yearID != 1987) %>% 
    filter(teamID != 'TEX')

salaries <- salaries %>%
  filter(1985 <= yearID & yearID <= 2016) %>%
  mutate(salaryMil = salary / 1000000) %>%
  filter(yearID != 1987) %>%  
    filter(teamID != 'TEX')

#Clearly missing lots of data for this datapoint hence its removal https://www.baseball-reference.com/teams/TEX/1987.shtml
```

```{r compute-team-salary-total, message = FALSE}
teams <- teams %>%
  inner_join(salaries) %>%
  group_by(yearID, teamID, G, W, L, WSWin, winPercentage) %>%
  summarize(totalSalaryMil = sum(salaryMil))
```

```{r compute-player-salary-share, message = FALSE}
salaries <- salaries %>% 
  inner_join(teams) %>%
  mutate(salaryShare = salaryMil / totalSalaryMil * 100) %>%
  mutate(salaryShareSquared = salaryShare ^ 2) %>%
  select(yearID, teamID, playerID, salary, salaryShare, salaryShareSquared)
```

```{r compute-team-salary-hhi, message = FALSE}
teams <- teams %>%
  inner_join(salaries) %>%
  group_by(yearID, teamID, G, W, L, winPercentage, WSWin, totalSalaryMil) %>%
  summarize(HHI = sum(salaryShareSquared))
```

```{r subset-old-data, message = FALSE}
teams_old <- teams %>%
  filter(1985 <= yearID & yearID <= 1998) %>%
  mutate(normalizedYear = yearID - 1985)

salaries_old <- salaries %>%
  filter(1985 <= yearID & yearID <= 1998)
```

```{r summary-statistics}
summary(teams_old$winPercentage)
sd(teams_old$winPercentage)
summary(teams_old$totalSalaryMil)
sd(teams_old$totalSalaryMil)
summary(teams_old$HHI)
sd(teams_old$HHI)
```

```{r linear-fixed-effects-regression-old}
linear_fixed_old <- lm(formula = winPercentage ~ totalSalaryMil + HHI +
                                                 normalizedYear + teamID + 0,
                       data = teams_old)
summary(linear_fixed_old)$coefficients[1:3,]
```

```{r linear-random-effects-regression-old}
linear_random_old <- lm(formula = winPercentage ~ totalSalaryMil + HHI + normalizedYear,
                        data = teams_old)
summary(linear_random_old)$coefficients[1:4,]
```
## Reproduction Notes

- original author did not describe how time fixed effects are accounted for (across expansion periods or every year)
- no discussion about limiting to 25 man roster vs 40 man roster
- no discussion of cut players, traded players
- no discussion of signing bonuses

## Reproduction Analysis


# Extension

## Extension Code

```{r salary-vs-time, message = FALSE, out.width = '70%', fig.align = 'center'}
salary_vs_time <- salaries %>%
  group_by(yearID) %>%
  summarize(avg = mean(salary), max = max(salary))

ggplot(data = salary_vs_time) +
  geom_point(aes(x = yearID, y = avg, color = 'Average')) +
  geom_smooth(aes(x = yearID, y = avg, color = 'Average')) + 
  geom_point(aes(x = yearID, y = max, color = 'Maximum')) + 
  geom_smooth(aes(x = yearID, y = max, color = 'Maximum')) +
  scale_color_manual(values = c('#4286f4', '#f44741')) +
  scale_y_log10(labels = comma) + 
  labs(color = 'Type') +
  xlab('Year') +
  ylab('Player Salary')
```

```{r player-salary-distributions, message = FALSE, out.width = '55%', fig.align = 'center'}
salaries_1985 <- filter(salaries, yearID == 1985)
salaries_1998 <- filter(salaries, yearID == 1998)
salaries_2016 <- filter(salaries, yearID == 2016)

ggplot(data = salaries_1985) +
  geom_histogram(aes(x = salary, y = (..count..) / sum(..count..))) +
  scale_x_continuous(labels = comma) + 
  xlab('Player Salary in 1985') +
  ylab('Fraction of All Salaries in 1985')

ggplot(data = salaries_1998) +
  geom_histogram(aes(x = salary, y = (..count..) / sum(..count..))) +
  scale_x_continuous(labels = comma) + 
  xlab('Player Salary in 1998') +
  ylab('Fraction of All Salaries in 1998')

ggplot(data = salaries_2016) +
  geom_histogram(aes(x = salary, y = (..count..) / sum(..count..))) +
  scale_x_continuous(labels = comma) + 
  xlab('Player Salary in 2016') +
  ylab('Fraction of All Salaries in 2016')
```

```{r team-salary-vs-time, message = FALSE, out.width = '80%', fig.align = 'center'}
current_teamIDs <- c('ARI', 'ATL', 'BAL', 'BOS', 'CHA', 'CHN', 'CIN', 'CLE', 'COL', 'DET',
                     'HOU', 'KCA', 'LAA', 'LAN', 'MIA', 'MIL', 'MIN', 'NYA', 'NYN', 'OAK',
                     'PHI', 'PIT', 'SDN', 'SEA', 'SFN', 'SLN', 'TBA', 'TEX', 'TOR', 'WAS')
team_colors <- c('#cccccc', '#cccccc', '#cccccc', '#BD3039', '#cccccc',
                 '#cccccc', '#cccccc', '#cccccc', '#cccccc', '#cccccc',
                 '#cccccc', '#cccccc', '#cccccc', '#0157a8', '#cccccc',
                 '#cccccc', '#cccccc', '#11325b', '#cccccc', '#04683b',
                 '#cccccc', '#cccccc', '#cccccc', '#cccccc', '#cccccc',
                 '#cccccc', '#cccccc', '#cccccc', '#cccccc', '#cccccc')
colored_teamIDs <- c('BOS', 'LAN', 'NYA', 'OAK')

team_salary_vs_time <- salaries %>%
  filter(teamID %in% current_teamIDs) %>%
  group_by(yearID, teamID) %>%
  summarize(avg = mean(salary)) %>%
  mutate(flag = teamID %in% colored_teamIDs)

underlay_data <- filter(team_salary_vs_time, !flag)
overlay_data <- filter(team_salary_vs_time, flag)

ggplot() +
  geom_point(data = underlay_data, aes(x = yearID, y = avg, color = teamID)) +
  geom_line(data = underlay_data, aes(x = yearID, y = avg, color = teamID)) +
  geom_point(data = overlay_data, aes(x = yearID, y = avg, color = teamID)) +
  geom_line(data = overlay_data, aes(x = yearID, y = avg, color = teamID)) +
  scale_y_continuous(labels = comma) + 
  scale_color_manual(values = team_colors) +
  labs(color = 'Team') +
  xlab('Year') +
  ylab('Average Team Salary')
```

```{r hhi-vs-time, message = FALSE, out.width = '70%', fig.align = 'center'}
hhi_vs_time <- teams %>%
  group_by(yearID) %>%
  summarize(avg = mean(HHI))

ggplot(data = hhi_vs_time) +
  geom_point(aes(x = yearID, y = avg)) +
  geom_smooth(aes(x = yearID, y = avg)) + 
  xlab('Year') +
  ylab('Average Team HHI')
```

```{r hhi-distributions, message = FALSE, out.width = '55%', fig.align = 'center'}
ggplot(data = filter(teams, mean(teams$HHI) - 5 * sd(teams$HHI) <= HHI & HHI <= mean(teams$HHI) + 5 * sd(teams$HHI))) +
  geom_histogram(aes(x = HHI, y = (..count..) / sum(..count..))) +
  geom_vline(data = filter(teams, WSWin), aes(xintercept = HHI), color = 'blue', linetype = 'dashed') +
  xlab('Team HHI') +
  ylab('Fraction of All HHI Values')
```

```{r hhi-vs-total-salary, message = FALSE, out.width = '70%', fig.align = 'center'}
year_to_period <- function(year) {
  if (year <= 1992) {
    return('Pre 1993')
  }
  else if (1993 <= year & year <= 1997) {
    return('Between 1993 and 1997')
  }
  else {
    return('Post 1997')
  }
}

hhi_vs_total_salary <- teams %>%
  mutate(period = year_to_period(yearID))
hhi_vs_total_salary$period <- factor(hhi_vs_total_salary$period, levels = c('Pre 1993', 'Between 1993 and 1997', 'Post 1997'))

ggplot(data = hhi_vs_total_salary) +
  geom_point(aes(x = totalSalaryMil * 1000000, y = HHI, color = period)) +
  scale_x_log10(labels = comma) +
  scale_y_log10() +
  labs(color = 'Period') +
  xlab('Total Team Salary') +
  ylab('Team HHI')
```

```{r subset-new-data, message = FALSE}
teams_new <- teams %>%
  filter(1999 <= yearID & yearID <= 2016) %>%
  mutate(normalizedYear = yearID - 1999)

salaries_new <- salaries %>%
  filter(1999 <= yearID & yearID <= 2016)
```

```{r linear-fixed-effects-regression-new}
linear_fixed_new <- lm(formula = winPercentage ~ totalSalaryMil + HHI +
                                                 normalizedYear + teamID + 0,
                       data = teams_new)
summary(linear_fixed_new)$coefficients[1:3,]
```

```{r linear-random-effects-regression-new}
linear_random_new <- lm(formula = winPercentage ~ totalSalaryMil + HHI + normalizedYear,
                        data = teams_new)
summary(linear_random_new)$coefficients[1:4,]
```
```{r compute new inequality index- gini- and run regressions}
#Gini coef: https://en.wikipedia.org/wiki/Gini_coefficient

gini_coef <- salaries %>%
  group_by(yearID, teamID) %>%
    summarize(gini_coef = Gini(salary))

teams <- teams %>%
  inner_join(gini_coef)

#Compare Gini index to HHI for old teams
gini_old <- teams %>%
  filter(1985 <= yearID & yearID <= 1998) %>%
  mutate(normalizedYear = yearID - 1985)

gini_fixed_old <- lm(formula = winPercentage ~ totalSalaryMil + gini_coef +
                                                 normalizedYear + teamID + 0,
                       data = gini_old)
summary(gini_fixed_old)$coefficients[1:3,]

gini_random_old <- lm(formula = winPercentage ~ totalSalaryMil + gini_coef + normalizedYear,
                        data = gini_old)
summary(gini_random_old)$coefficients[1:4,]


#Compare Gini coef to HHI results for new teams
gini_new <- teams %>%
  filter(1999 <= yearID & yearID <= 2016) %>%
  mutate(normalizedYear = yearID - 1999)

gini_fixed_new <- lm(formula = winPercentage ~ totalSalaryMil + gini_coef +
                                                 normalizedYear + teamID + 0,
                       data = gini_new)
summary(gini_fixed_new)$coefficients[1:3,]

gini_random_new <- lm(formula = winPercentage ~ totalSalaryMil + gini_coef + normalizedYear,
                        data = gini_new)
summary(gini_random_new)$coefficients[1:4,]
```
```{r compute new inequality index- Atkinson index- and run regressions}
#Atkinson index: https://en.wikipedia.org/wiki/Atkinson_index
atkinson <- salaries %>%
  group_by(yearID, teamID) %>%
    summarize(atk = Atkinson(salary))

teams <- teams %>%
  inner_join(atkinson)

#Compare Atkinson index to HHI for old teams
atk_old <- teams %>%
  filter(1985 <= yearID & yearID <= 1998) %>%
  mutate(normalizedYear = yearID - 1985)

atk_fixed_old <- lm(formula = winPercentage ~ totalSalaryMil + atk +
                                                 normalizedYear + teamID + 0,
                       data = atk_old)
summary(atk_fixed_old)$coefficients[1:3,]

atk_random_old <- lm(formula = winPercentage ~ totalSalaryMil + atk + normalizedYear,
                        data = atk_old)
summary(atk_random_old)$coefficients[1:4,]


#Compare Atkinson coef to HHI results for new teams
atk_new <- teams %>%
  filter(1999 <= yearID & yearID <= 2016) %>%
  mutate(normalizedYear = yearID - 1999)

atk_fixed_new <- lm(formula = winPercentage ~ totalSalaryMil + atk +
                                                 normalizedYear + teamID + 0,
                       data = atk_new)
summary(atk_fixed_new)$coefficients[1:3,]

atk_random_new <- lm(formula = winPercentage ~ totalSalaryMil + atk + normalizedYear,
                        data = atk_new)
summary(atk_random_new)$coefficients[1:4,]
```

```{r k-fold model predict, not splitting on time}
teams <- teams %>%
  mutate(normalizedYear = yearID - 1985) %>% 
  ungroup()

set.seed(42)
num_folds <- 5
num_rows <- nrow(teams)
ndx <- sample(1:num_rows, num_rows, replace=F)

train_data <- teams[ndx, ] %>%
  mutate(fold = (row_number() %% num_folds) + 1)

validate_err <- c()
train_err <- c()
for (f in 1:num_folds) {
  temp_train <- filter(train_data, fold != f)
  
  #Using fixed effects model because the data is randomly sorted
  model <- lm(formula = winPercentage ~ totalSalaryMil + HHI +
                                                 normalizedYear + teamID + 0, data = temp_train)
  train_err[f] <- sqrt(mean( (predict (model, temp_train) - temp_train$winPercentage) ^2) )
  
  # evaluate on the validation data
  temp_validate <- filter(train_data, fold == f)
  validate_err[f] <- sqrt(mean( (predict (model, temp_validate) - temp_validate$winPercentage) ^2) )
}

# compute the average validation error across folds and the standard error on this estimate
avg_validate_err <- mean(validate_err)
se_validate_err <- sd(validate_err) / sqrt(num_folds)

avg_train_err <- mean(train_err)
se_train_err <- sd(train_err) / sqrt(num_folds)
```

```{r use old years to train model predict on new years}
time_train_data <- teams %>% filter(yearID < 2012)
time_validate_data <- teams %>% filter(yearID>2011)

#Using random effects model because we dont want to worry about team being good in 80s/90s and bad in 2010s
model <- lm(formula = winPercentage ~ totalSalaryMil + HHI +
              normalizedYear,
            data = time_train_data)

time_train_err <- sqrt(mean( (predict (model, time_train_data) - time_train_data$winPercentage) ^2) )
  
time_validate_err <- sqrt(mean( (predict (model, time_validate_data) - time_validate_data$winPercentage) ^2) )
```

## Extension Notes
- note that minimum salary has increased over time: https://www.baseball-reference.com/bullpen/Minimum_salary

- Coefficients of different inequality indexes are the same sign and have the same predictive impact (magnitudes are different because HHI has a wider range than the other two which are between 0 and 1)

- Used fixed effects model when split data independent of time, ie teams from 1985 and 2016 are in the training set. Meanwhile used random effects model when split data based on time, ie teams from 1985-2011 were in training set and 2012-2016 were test set. The use of the random effects model for the latter was to account for the fact that teams might be dominant in early years but not so much in more recent years. Also handles the issue of teams not existing in training set but potentially in validation set.

## Extension Analysis
- 6.4%-6.6% error in win percentage predictions for both kinds of regressions. This isn't amazing given that the range of reasonable values is roughly 40%-70% win percentage

# Postface

The following is a list of all packages used to generate these results.

```{r}
sessionInfo()
```